Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-12-11 paper.tex
Add explicit reference
Reviewer B:
  Line 399: "Our previous work on Usuba suggests ..." is there a reference for this work?
Tone down banter
Reviewer B:
  Footnote 3: not useful.

The type \coqe!bool! is an obvious instance of a Boolean algebra.
However, any commutative applicative~\citep{mcbride:applicative}
functor \coqe!F! applied to a Boolean algebra also yields a Boolean
algebra (dispatching the Boolean operations pointwise to the
underlying elements). For the sake of completeness, we recall that an
applicative functor offers the following operations:
%
\coqfrom{Algebra}{Applicative}

@@

The confusion layer builds upon two permutations over \coqe!Slice8.T!.
Once again, this is merely exploiting the Naperian structure of
\coqe!Slice8.T!: for any Naperian functor \coqe!F! indexed by
\coqe!Ix!, there exists a generic permutation operator
%
\coqe!perm : forall A, (Ix -> Ix) -> F A -> F A!.
%
This leads to the following definitions:
%
\coqfrom{Skinny/Base}{perms}

@@

Storing only 16 bits in a 32 bits word (on \textsc{Arm} Cortex M)
remains sub-optimal. We can further increase register usage and
instantly double throughput by processing two blocks at the same
time. To this end, we introduce a new Naperian functor,
\coqe!Double.T!, indexed by the two elements set \coqe!Fst!  and
\coqe!Snd!. This leads to
%
\coqfrom{Skinny/Bitslice}{state}
%
which denote our intent to manipulate objects of type \coqe!reg32! as
if they were stored in a single machine word. In the present work,
this observation will remain at the state of wishful thinking. One can
think of the code in this section (as well as the fixsliced code, in
the next section) as playing, themselves, the role of specifications
to lower-level refinements. Our previous work on Usuba suggests that
we will be able to deliver on this promise in the future. Further,
remark that the \coqe!state! now corresponds to two intermingled
blocks but as we shall process them in lock-step throughout the
cipher, this does not make much difference.

@@

Indeed, aside from the logical operations, the rotations are in fact
almost free thanks to the \textsc{Arm} Cortex barrel
shifter.\footnote{By this remark, we hope to have established our
credentials as budding cryptographers.}

@@

The crux of the matter is the interaction between \coqe!phi_! and the
diffusion layer. We observe that putting a state \coqe!phi_ s! into
the diffusion layer first yield a transformation relativized to a
machine word followed by the emission of two \coqe!phi_! steps out of
the layer:
%
\begin{coro}
  \coqfrom{Skinny/FixsliceProof}{absorb_phi_sigma1}
\end{coro}

>>>>

The type \coqe!bool! is an obvious instance of a Boolean algebra.
However, any commutative applicative~\citep{mcbride:applicative}
functor \coqe!F! applied to a Boolean algebra also yields a Boolean
algebra (dispatching the Boolean operations pointwise to the
underlying elements). For the sake of completeness, we recall that an
applicative functor offers the following operations:
%
\coqfrom{Algebra}{Applicative}
%
subject to an equational theory expressing the fact that applicative
functors correspond to lax monoidal
functors~\citep{maclane:working-categorist} over the Cartesian product
type. Commutativity of an applicative functor \coqe!F! states that
arguments can be reduced in any order:
%
\coqfrom{Algebra}{ApplicativeComm}

@@

The confusion layer builds upon two permutations over \coqe!Slice8.T!.
Once again, this is merely exploiting the Naperian structure of
\coqe!Slice8.T!: for any Naperian functor \coqe!F! indexed by
\coqe!Ix!, we can shuffle the elements of a container \coqe!F A! following an arbitrary permutation of indices:
%
\coqe!perm : forall A, (Ix -> Ix) -> F A -> F A!.
%
This leads to the following definitions:
%
\coqfrom{Skinny/Base}{perms}

@@

Storing only 16 bits in a 32 bits word (on \textsc{Arm} Cortex M)
remains sub-optimal. We can further increase register usage and
instantly double throughput by processing two blocks at the same
time. To this end, we introduce a new Naperian functor,
\coqe!Double.T!, indexed by the two elements set \coqe!Fst!  and
\coqe!Snd!. This leads to
%
\coqfrom{Skinny/Bitslice}{state}
%
which denote our intent to manipulate objects of type \coqe!reg32! as
if they were stored in a single machine word. In the present work,
this observation will remain at the state of wishful thinking. One can
think of the code in this section (as well as the fixsliced code, in
the next section) as playing, themselves, the role of specifications
to lower-level refinements. Our previous work on Usuba~\citep{mercadier:PhD} suggests that
we will be able to deliver on this promise in the future. Further,
remark that the \coqe!state! now corresponds to two intermingled
blocks but as we shall process them in lock-step throughout the
cipher, this does not make much difference.

@@

Indeed, aside from the logical operations, the rotations are in fact
almost free thanks to the \textsc{Arm} Cortex barrel
shifter.

@@

The crux of the matter is the interaction between \coqe!phi_! and the
diffusion layer. We observe that putting a state \coqe!phi_ s! into
the diffusion layer first yields a transformation relativized to a
machine word followed by the emission of two \coqe!phi_! steps out of
the layer:
%
\begin{coro}
  \coqfrom{Skinny/FixsliceProof}{absorb_phi_sigma1}
\end{coro}
